White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2017-07-09 papers/wdmerger_II/paper.tex
Add a description of the hydro flux limiter

\shorttitle{WD Mergers. II. Collisions}
\shortauthors{Katz et al. (2016)}

@@

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI} (hereafter, Paper I).
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to couple them to ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

>>>>

\shorttitle{WD Mergers. II. Collisions}
\shortauthors{Katz et al. (2017)}

@@

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI} (hereafter, Paper I).
Revisions to the core algorithm since then are presented in \autoref{app:hydro}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to couple them to ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

\appendix

\section{Hydrodynamics Algorithm Updates}
\label{app:hydro}

The core \castro\ algorithm for hydrodynamic evolution is presented in
\citet{wdmergerI}, and is largely unchanged in the present work. This appendix
focuses on minor revisions to the algorithm that have occurred since then.

\subsection{Flux Limiting}
\label{app:hydro:flux_limiting}

Multi-dimensional hydrodynamic white dwarf simulations often have numerical artifacts
that result from the sharp surface between the white dwarf and the ambient material
in the simulation domain, which is difficult to resolve. A somewhat common issue, especially
at low resolution, is negative densities that occur as a result of a hydro update.
The \castro\ code contains a prescription for dealing with negative densities, that
resets the negative density to be similar to nearby zones. Various choices exist
for how to do this, such as resetting it to the original zone density before the
update or resetting it to some linear combination of the density of nearby zones.
The reset is problematic because the strategy is not unique and no choice is
clearly better than the rest in all cases. Additionally, it is not specified at all
how to reset momenta in such a case. Consequently, we desired to improve the situation
by limiting fluxes such that negative densities could not occur, so that such a
reset would in practice always be avoided. Our solution implements the positivity-preserving
method of \cite{hu:2013}. A hydrodynamical update to a zone can be broken down into
an update over every face of the zone where a flux crosses the face over the timestep.
The central insight of the positivity-preserving method is that if the update over every
face is positivity-preserving, then the total update must be positivity-preserving as well.
To guarantee positivity preservation at the zone edge ${\rm i}+1/2$, the flux
$\mathbf{F}^{n+1/2}_{{\rm i}+1/2}$ at that face is modified to become:
\begin{equation}
  \mathbf{F}^{n+1/2}_{{\rm i}+1/2} \rightarrow \theta_{{\rm i}+1/2} \mathbf{F}^{n+1/2}_{{\rm i}+1/2} + (1 - \theta_{{\rm i}+1/2}) \mathbf{F}^{LF}_{{\rm i}+1/2}, \label{eq:limited_flux}
\end{equation}
where $0 \leq \theta_{{\rm i}+1/2} \leq 1$ is a scalar, and $\mathbf{F}^{LF}_{{\rm i}+1/2}$ is the Lax-Friedrichs flux,
\begin{equation}
  \mathbf{F}^{LF}_{{\rm i}+1/2} = \frac{1}{2}\left[\mathbf{F}^{n}_{{\rm i}} + \mathbf{F}^{n}_{{\rm i}+1} + \text{CFL}\frac{\Delta x}{\Delta t} \frac{1}{\alpha}\left(\mathbf{U}^{n}_{{\rm i}} - \mathbf{U}^{n}_{{\rm i}+1}\right)\right],
\end{equation}
where $0 < \text{CFL} < 1$ is the CFL safety factor (the method is guaranteed to
preserve positivity as long as $\text{CFL} < 1/2$), and $\alpha$ is a scalar that
ensures multi-dimensional correctness ($\alpha = 1$ in 1D, $1/2$ in 2D, $1/3$ in 3D).
$\mathbf{F}_{{\rm i}}$ is the flux of material evaluated at the zone center ${\rm i}$
using the cell-centered quantities $\mathbf{U}$. The scalar $\theta_{{\rm i}+1/2}$ is
chosen at every interface by calculating the update that would be obtained from
\autoref{eq:limited_flux}, setting the density component equal to some small, positive
value, and solving for the value of $\theta$ at the interface that makes
the equality hold. In regions where the density is not at risk of going negative,
$\theta \approx 1$ and the original hydrodynamic update is recovered.
Further discussion, including a proof of the method, a description of multi-dimensional effects,
and test verification problems, can be found in \cite{hu:2013}.
